---
title: "Open Policing Project"
description: |
  Stopped Driver Age Comparison by City
author: Emmie Appl
date: April 30, 2025
format:
  html:
    toc: false
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
  message: false
bibliography: references.bib
---

The [Stanford Open Policing Project](https://openpolicing.stanford.edu/), published in @pierson2020, is a compilation of data acquired through millions of traffic stops made by law enforcement agencies across 42 different US states. Some of the data from these stops include records of the driver's age. I wanted to compare the average age of stopped drivers across different California cities, as well as how this metric changed over time. To this end, I analyzed the data for all California cities included in the Open Policing Project that recorded driver age, calculating the average age of stopped drivers in each city for each year.

```{r}
library(DBI)
con_traffic <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```

```{sql}
#| connection: con_traffic
#| output.var: "stop_age_table"

SELECT "San Francisco" as city,
  YEAR(date) AS stop_year,
  subject_age AS age
FROM ca_san_francisco_2020_04_01
WHERE subject_age IS NOT NULL
  AND date IS NOT NULL

UNION

SELECT "Bakersfield" as city,
  YEAR(date) AS stop_year,
  subject_age AS age
FROM ca_bakersfield_2020_04_01
WHERE subject_age IS NOT NULL
  AND date IS NOT NULL

UNION

SELECT "Long Beach" as city,
  YEAR(date) AS stop_year,
  subject_age AS age
FROM ca_long_beach_2020_04_01
WHERE subject_age IS NOT NULL
  AND date IS NOT NULL

UNION

SELECT "San Diego" as city,
  YEAR(date) AS stop_year,
  subject_age AS age
FROM ca_san_diego_2020_04_01
WHERE subject_age IS NOT NULL
  AND date IS NOT NULL

UNION

SELECT "Stockton" as city,
  YEAR(date) AS stop_year,
  subject_age AS age
FROM ca_stockton_2020_04_01
WHERE subject_age IS NOT NULL
  AND date IS NOT NULL

ORDER BY city, stop_year;
```

```{r}
dbDisconnect(con_traffic, shutdown = TRUE)
```

```{r}
library(dplyr)
stop_age_table <- stop_age_table |>
  mutate(
    age = as.numeric(age),
    stop_year = as.integer(stop_year)
  )
```

```{r}
every_three_years <- stop_age_table |>
  filter(stop_year %in% c(2008, 2011, 2014, 2017))

library(ggplot2)
ggplot(every_three_years, aes(x = city, y = age)) +
  facet_wrap(vars(stop_year)) +
  geom_boxplot() +
  scale_x_discrete(labels = c(
    "San Diego" = "SD",
    "San Francisco" = "SF",
    "Long Beach" = "LB"
  )) +
  labs(title = "Distribution of Average Stopped Driver Age by City Every Three Years",
       x = "City",
       y = "Stopped Driver Age")
```

This plot is a faceted box plot comparing the stopped driver ages between California cities in the years 2008, 2011, 2014, and 2017. The x-axis represents the 5 different cities (L to R: Bakersfield, Long Beach (LB), San Diego (SD), San Francisco (SF), Stockton) and the y-axis shows the stopped driver age. In each boxplot, the vertical spread represents the variation in the average stopped driver age across different years. The middle line of each box marks the median age, and the box boundaries represent the interquartile range. Individual points represent the individual stopped driver ages from a singular year. The absence of a boxplot represents the absence of data for that city in that year. San Francisco consistently had a higher median stopped driver age compared to the other California cities in all years that data was collected. Most cities had a median stropped driver age of or slightly above 50 years.

```{r}
avg_ages <- stop_age_table |>
  group_by(city, stop_year) |>
  summarize(avg_age = mean(age), .groups = "drop")

avg_ages |>
  ggplot(aes(x = stop_year, y = avg_age)) +
  scale_x_continuous(breaks = seq(2007, 2018, by = 1)) +
  geom_line(aes(color = city)) +
  labs(
    x = "Year",
    y = "Average Stopped Driver Age",
    color = "City",
    title = "Average Age of Stopped Drivers in California Cities"
  )
  
```

This plot is a line plot tracking the average stopped driver age each year in each of the 5 cities. The x-axis represents the year that the stop was made and the y-axis shows the average age of drivers stopped in that year. Each line represents a different city and connects the yearly averages to illustrate trends in driver age over time. The average age of stopped drivers in San Francisco dipped from about 56 to 53 after 2008 and began climbing again starting in 2011. Bakersfield saw a steady increase of about 2 years between 2008 and 2010, then a drop to an average age of about 35.5 years that remained consistent until the last data point from 2018. The average stopped driver age in Long Beach also rose by approximately 2 years between 2008 and 2013, remaining stable until 2016 when there was sharp decrease. Although the data for Stockton and San Diego only cover 4 and 3 years, respectively, the average stopped driver ages remained within about a 2 year window.

## **Conclusions**

Together, these two plots compare the average age of stopped drivers between California cities, as well as how that number has changed over time. Comparing the average age of stopped drivers to the average age of all drivers in these cities will lead to understanding whether certain age ranges are more prone to being stopped by law enforcement, or whether these ages simply reflect the average age of the driver population in that city. Seeing how the average stopped driver ages change over time compared to changes in the average driver population age, as well as the average general population age, will reveal whether the changes in average stopped driver age are due to changes in city demographics, or if there may be other underlying reasons for these changes.
